Analysis of observational data, to calculate the Delayed-Tau model parameters

### 2025-01-15 thesis.qmd
Add thesis title

---
title: "thesis"
author: "Dimitris Papachristopoulos"
format: 
  pdf: default
  #logo: ./figure/
  arxiv-pdf:
    keep-tex: true    
    link-citations: true
    link-bibliography: true
toc: true
toc-depth: 3
theme: minti
bibliography: ./My_Library.bib
execute: 
  echo: false
  eval: true
  warning: false
editor: visual
---

@@

@haslbauerCosmologicalStarFormation2023 argue that the Delayed-$\tau$ model is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses [^1] from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LCV), allows the investigation of the SFR throughout the life each galaxy and so we can find the expected time of peak of the SFR.

[^1]: They calculated the Stellar Masses by using a Mass to Light ratio of 0.6

@@

The study of galaxies is a very active field of astronomy since it is a relatively young discipline. Until 1920, astronomers who observed spiral nebulae were not certain what they were. In 1921, two papers were published: one argued that the Milky Way constituted the whole Universe and that the spiral nebulae were part of it, while the second argued that each spiral nebula was, in fact, a distinct "island universe" [@shapleyScaleUniverse1921]. This debate became known as the Great Debate[^2].

[^2]: @hoskinGreatDebateWhat1976

@@

The most common method of classification is the *Hubble-de Vacouleur* classification, which categorizes galaxies based on their morphology.

@@

-   **Lenticular Galaxies (S0):** Intermediate between elliptical and spiral galaxies,

@@

Intermediate galaxies between ellipticals and spirals are called **Lenticular galaxies (S0-SB0)**, featuring a central bulge and disk but lacking significant spiral structure.

[![Tuning-fork-style diagram of the Hubble sequenceBy Cosmogoblin - Own work, CC0, https://commons.wikimedia.org/w/index.php?curid=121743256](figure/1311px-Hubble_Tuning_Fork_diagram.svg.png){fig-align="center" width="400"}](https://commons.wikimedia.org/w/index.php?curid=121743256)

@@

@@

@@

SFR can be estimated using various photometric or spectroscopic methods based on the luminosity of at least one spectral band or the intensity of a spectral line. Different luminosities and intensities trace distinct emission mechanisms, offering insights into a galaxy’s radiation sources. Below are common methods:[^3]

[^3]: @calzettiStarFormationRate2012, @mushotzkyASTR620GalaxiesFall2017

@@

```{r}
#| fig.cap="Star Formation Rate over the time of star formation, for different parametric models"
library(ggplot2)
library(dplyr)
library(latex2exp)

# Define time values (positive times)
t <- seq(0.01, 12, length.out = 200)

# Define SFR models
exponential_decline <- function(t, tau) exp(-t / tau)
delayed_exponential <- function(t, tau) t * exp(-t / tau)
log_normal <- function(t, tau, T0) (1 / tau) * exp(-((log(t) - log(T0))^2) / (2 * tau^2))
double_power_law <- function(t, tau, alpha, beta) 1 / ((t / tau)^alpha + (t / tau)^(-beta))

# Generate data for each model
data <- data.frame(
  t = rep(t, 4),
  SFR = c(
    exponential_decline(t, tau = 2),
    delayed_exponential(t, tau = 2),
    log_normal(t, tau = 2, T0 = 10),
    double_power_law(t, tau = 2, alpha = 2, beta = 5)
  ),
  Model = rep(c("\nExponential Decline\n(τ=2 Gyr)", "\nDelayed Exponential\n(τ=2 Gyr)", "\nLog-Normal\n(τ = 2 Gyr, To = 10 Gyr)", "\nDouble Power Law\n(τ = 2 Gyr, α = 2, β = 5)"), each = length(t))
)

# Plot all models
ggplot(data, aes(x = t, y = SFR, color = Model)) +
  geom_line(size = 1) +
  labs(
    title = "Star Formation Rate Models",
    x = TeX("Star Formation time ( $t_{sf}$ ) [Gyr]"),
    y = TeX("SFR($t_{sf}$) [ M$_*$/yr ]" ),
    color = "Model"
  ) +
  scale_x_continuous(limits = c(0, 12)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

@@

$V = H_0\times d$, where $H_0 \approx 69.8 \text{ km/s/Mpc}$ is the Hubble constant and $d$ is the distance between the two galaxies.[^4]

[^4]: We also have non-Hubble motions $V = H_0\times d + V_0$, where $V_0$ is the peculiar velocity and it could, for example, be due to galaxy cluster dynamics. For the current explanation we are going to ignore it. This way the radial velocity $v$ is equal to $V$

@@

@@

>>>>

---
title: "Analysis of observational data, to calculate the Delayed-Tau model parameters"
author: "Dimitrios Papachristopoulos"
format: 
  pdf: default  
  # arxiv-pdf:
  #    keep-tex: true    
  #    link-citations: true
  #    link-bibliography: true
  #    runninghead: ""
toc: true
toc-depth: 3
bibliography: ./My_Library.bib
execute: 
  echo: false
  eval: true
  warning: false
editor: visual
---

@@

@haslbauerCosmologicalStarFormation2023 argue that the Delayed-$\tau$ model is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LCV), allows the investigation of the SFR throughout the life each galaxy and so we can find the expected time of peak of the SFR.

@@

The study of galaxies is a very active field of astronomy since it is a relatively young discipline. Until 1920, astronomers who observed spiral nebulae were not certain what they were. In 1921, two papers were published: one argued that the Milky Way constituted the whole Universe and that the spiral nebulae were part of it, while the second argued that each spiral nebula was, in fact, a distinct "island universe" [@shapleyScaleUniverse1921]. This debate became known as the Great Debate[^1].

[^1]: @hoskinGreatDebateWhat1976

@@

One of the most common methods of classification is the *Hubble classification [@hubbleCepheidsSpiralNebulae1925]* , which categorizes galaxies based on their morphology.

@@

-   **Lenticular Galaxies (S0):** Intermediate between elliptical and spiral galaxies, featuring a central bulge and disk but lacking significant spiral structure.

@@

::: {#fig-morphology layout-ncol="2"}
![Tuning-fork-style diagram of the Hubble sequence By Cosmogoblin - Own work, CC0, <https://commons.wikimedia.org/w/index.php?curid=121743256>](figure/1311px-Hubble_Tuning_Fork_diagram.svg.png){#fig-hubble}

![Hubble-de Vaucouleurs classification system By @devaucouleursClassificationMorphologyExternal1959](figure/de-vacouler-system.png){#fig-deVac}

Diagrams visualizing the Hubble and Hubble-de Vaucouleurs morphological classification systems
:::

Although Hubble’s original scheme was revolutionary, **de Vaucouleurs** recognized that not all galaxies fit neatly into its categories. In response, he introduced a more nuanced classification system [@devaucouleursClassificationMorphologyExternal1959] that:

-   **Accounts for Rings**: Some galaxies feature ring-like structures around their bulge or bar. To denote this, de Vaucouleurs added **(R)** to the classification (for example, (R)SBa).

-   **Differentiates Bar Strength**: Since bars can vary from subtle to dominant, he proposed **SA** (unbarred), **SAB** (weakly barred), and **SB** (strongly barred).

-   **Incorporates Numerical Types (T-Types)**: To capture subtle transitions along the morphological sequence, de Vaucouleurs assigned a numerical index (TTT) ranging from -6 (pure compact ellipticals) to +10 (extreme irregulars). Intermediate values (e.g., -1 for S0, 2 for Sab, 5 for Sc, etc.) let astronomers pinpoint galaxies that don’t fit cleanly into the original categories.

This expanded framework also embraces **early-type** and **late-type** galaxies as part of a continuous evolutionary sequence:

-   **Early-type galaxies** (E and S0):

    -   Smooth appearance.

    -   Predominantly older stellar populations.

    -   Minimal amounts of gas and dust.

    -   Assigned negative T-values ($-6 \le T < 0$).

-   **Late-type galaxies** (Spirals and Irregulars):

    -   Rich in gas and dust.

    -   Significant ongoing star formation.

    -   Smaller bulges and more open arms from Sa/SBa to Sc/SBc.

    -   Assigned positive T-values , extending to +11 for extreme irregulars.

By offering extra designations for bar strength, ring features, and transitional morphologies, de Vaucouleurs’ system paints a more complete picture of galaxies and how they evolve. It allows researchers to quantify where a galaxy lies along the continuum, rather than forcing it into a single rigid label.

@@

Despite the usefulness of de Vaucouleurs’ numerical T-type system , we run into practical issues when classifying dwarf galaxies. These galaxies need a more detailed classification system, since the T-Type system can denote galaxies, whose physical properties drastically differ, with by the same number. For instance, dwarf spheroidals and normal ellipticals both end up with a T-value below zero (T \< 0), even though their properties are vastly different. Then there are “transient” dwarf galaxies (Tr), whose features bridge spheroidal (Sph)(Sph) and irregular (Ir)(Ir) types. Inaccuracies in classification can cause these hybrids to “jump” from one extreme of the T-scale to the other with only a small error in morphology.

To address this, van den Bergh suggested dividing dwarf galaxies by luminosity class, which prevents them from leaping between extreme categories. This refined approach better reflects the intrinsic diversity of dwarf systems, ensuring that subtle morphological differences are more accurately captured and reducing the risk of placing galaxies at the wrong end of the classification scale [@karachentsevUPDATEDNEARBYGALAXY2013]

@@

One of the main ingredients of the galaxies are large scale molecular clouds, rich in hydrogen, with masses of order $10^5\ M_\odot$, typical dimensions of $\sim 10$ parsec, temperatures of 10-100 K and densities of 10-300 molecules/cm$^3$ [@polsStellarStructureEvolution2009]. Stars are created inside these clouds when a perturbation disturbs them, and thus their pressure equilibrium, and they start to collapse into smaller clouds under their self-gravity.

The collapse leads to the formation of protostars, which eventually become main-sequence stars. The presence of dust within these clouds is crucial, as it shields the interior regions from ultraviolet radiation, allowing the gas to cool to temperatures below 100 K, facilitating star formation .

The location of the star forming regions within a galaxy depend on the morphology. In spiral galaxies, star formation mainly occurs along the disk, where the molecular clouds are dense due to the compression caused by spiral density waves. The compression not only initiates the collapse of the clouds and thus the star formation, but also feeds the arms with gas, which sustains the star formation.

In elliptical galaxies, on the other hand, star formation is minimal because they lack the cold gas reservoirs needed for new stars to form. Irregular galaxies, with their chaotic structures, often have patchy but vigorous star-forming regions, as they retain significant amounts of gas.

A special case are the starburst galaxies, who have a extremely active star forming regions, and seem to convert the gas into stars extremely fast (even 100 times faster than the Milky Way). These starbursts often concentrate their intense star-formation activity in compact regions about 1 kpc in size (typically in galaxy nuclei). Due to their high star-formation rates, starbursts host large numbers of young stars.

@@

SFR can be estimated using various photometric or spectroscopic methods based on the luminosity of at least one spectral band or the intensity of a spectral line. Different luminosities and intensities trace distinct emission mechanisms, offering insights into a galaxy’s radiation sources. Below are common methods:[^2]

[^2]: @calzettiStarFormationRate2012, @mushotzkyASTR620GalaxiesFall2017

@@

![Star Formation Rate over the time of star formation, for different parametric models](figure/sfr-models.png){fig-align="center" width="350"}

@@

$V = H_0\times d$, where $H_0 \approx 69.8 \text{ km/s/Mpc}$ is the Hubble constant and $d$ is the distance between the two galaxies.[^3]

[^3]: We also have non-Hubble motions $V = H_0\times d + V_0$, where $V_0$ is the peculiar velocity and it could, for example, be due to galaxy cluster dynamics. For the current explanation we are going to ignore it. This way the radial velocity $v$ is equal to $V$

@@

![](figure/lily-madau.png){fig-align="center" width="450"}

@@

More details about the model and how it clashes with the Lilly-Madau plot
